Procurements of the Canadian Government, Runaway Military Contracts

### 2024-12-03 paper/paper.qmd
added diagnostics subsection + diagnostics tables in appendix

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(zoo)
library(broom)

@@

# Create the bar chart
ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = def_c) +
  labs(x = "Year", y = "Total Spending (in Billions)") +
  theme_minimal() +
  scale_y_continuous(labels = dollar_format(prefix = "$")) +
  theme(plot.title = element_text(hjust = 0.5))

@@

# Create the bar chart
ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = public_c) +
  labs(x = "Year", y = "Total Spending (in Billions)") +
  theme_minimal() +
  scale_y_continuous(labels = dollar_format(prefix = "$")) +
  theme(plot.title = element_text(hjust = 0.5))

@@

# Create the grouped bar chart
ggplot(data = grouped_df, aes(x = year, y = total_spending_billion, fill = buyer_category)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9), # Adjust dodge width
    width = 0.8 # Adjust bar width
  ) +
  scale_fill_manual(
    values = c(
  "National Defence" = def_c,
  "Public Services and Procurement Canada" = public_c,
  "Shared Services Canada" = services_c,
  "Public Health Agency of Canada" = health_c,
  "Fisheries and Oceans Canada" = fish_c,
  "Other"  = "gray"
    ),
  guide = guide_legend(nrow = 3)
  ) +
  scale_y_continuous(
    labels = comma_format(accuracy = 1), # Formats y-axis labels without decimal places
    breaks = pretty_breaks(n = 5), # Ensures approximately 5 breaks like 5, 10, 15
    expand = expansion(mult = c(0, 0.05)) # Adds a small space above the highest bar
  ) +
  labs(
    x = "Year",
    y = "Total Spending (Billions)",
    fill = "Buyer",
    caption = "."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom"
  )

@@

```{r fig-distibution}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

@@

print(paste0("Percentage of contracts <= 100,000 CAD: ", round(percent_less_than * 100, 2), "%"))

@@

# Step 6: Create the ggplot2 bar chart
ggplot(contract_counts_df, aes(x = amount_label, y = Contract_Count_k)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  scale_y_continuous(
    labels = if (max_count >= 1000) {
      label_number(scale = 1, accuracy = 1)
    } else {
      label_number(accuracy = 1)
    },
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Number of Contracts per 10,000 CAD Award Amount Range",
    x = "Award Amount (Thousands)",
    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else ""),
    caption = "Data includes contracts <= 100,000 CAD."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
  )
```

@@

```{r fig-small-big}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

# Step 1: Create a new column 'contract_size' categorizing contracts
df <- df %>%
  mutate(
    contract_size = ifelse(amount <= 100000, "Small Contracts (≤ 100k)", "Big Contracts (> 100k)")
  )

@@

# Create the bar chart
ggplot(total_spent_df, aes(x = contract_size, y = total_amount_billion, fill = contract_size)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(round(total_amount_billion, 2), "B")), 
            vjust = -0.5, 
            size = 5) +  # Adds labels above the bars
  scale_fill_manual(
    values = c("Small Contracts (≤ 100k)" = "steelblue", "Big Contracts (> 100k)" = "tomato")
  ) +
  scale_y_continuous(
    labels = scales::number_format(scale = 1, suffix = "B"),  # Displays y-axis in billions
    expand = expansion(mult = c(0, 0.1))  # Adds space above the bars for labels
  ) +
  labs(
    title = "Total Amount Spent on Contracts",
    subtitle = "Comparison between Small (≤ 100k CAD) and Big (> 100k CAD) Contracts",
    x = "Contract Size",
    y = "Total Amount Spent (Billions of CAD)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    legend.position = "none",  # Removes the legend as it's redundant
    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")
  )

@@

@@

TODO Add Diagnostics

>>>>

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(zoo)
library(broom)
library(car)
library(caret)

@@

# Create the bar chart
ggplot(data = military_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = def_c) +
  labs(x = "Year", y = "Total Spending (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

# Create the bar chart
ggplot(data = public_spending_per_year, aes(x = year, y = total_spending)) +
  geom_bar(stat = "identity", fill = public_c) +
  labs(x = "Year", y = "Total Spending (Billions)") +
  theme_minimal() +
  scale_y_continuous() +
  theme(plot.title = element_text(hjust = 0.5))

@@

# Create the grouped bar chart
ggplot(data = grouped_df, aes(x = year, y = total_spending_billion, fill = buyer_category)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9), # Adjust dodge width
    width = 0.8 # Adjust bar width
  ) +
  scale_fill_manual(
    values = c(
  "National Defence" = def_c,
  "Public Services and Procurement Canada" = public_c,
  "Shared Services Canada" = services_c,
  "Public Health Agency of Canada" = health_c,
  "Fisheries and Oceans Canada" = fish_c,
  "Other"  = "gray"
    ),
  guide = guide_legend(nrow = 3)
  ) +
  scale_y_continuous(
    labels = comma_format(accuracy = 1),
    breaks = pretty_breaks(n = 5),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Year",
    y = "Total Spending (Billions)",
    fill = "Buyer",
    caption = "."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom"
  )

@@

```{r fig-distibution}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of Contracts that are less than $100,000
#| fig-align: center

@@

@@

# Step 6: Create the ggplot2 bar chart
ggplot(contract_counts_df, aes(x = amount_label, y = Contract_Count_k)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  scale_y_continuous(
    labels = if (max_count >= 1000) {
      label_number(scale = 1, accuracy = 1)
    } else {
      label_number(accuracy = 1)
    },
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Award Amount (Thousands)",
    y = paste0("Number of Contracts ", if (max_count >= 1000) "(Thousands)" else "")
  ) +
  theme_minimal()
```

@@

```{r fig-small-big}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Total Amount Spent on Contracts
#| fig-align: center

# Step 1: Create a new column 'contract_size' categorizing contracts
df <- df %>%
  mutate(
    contract_size = factor(
      ifelse(amount <= 100000, 
             "Small Contracts (≤ $100,000)", 
             "Big Contracts (> $100,000)"),
      levels = c("Small Contracts (≤ $100,000)", "Big Contracts (> $100,000)")  # Specify order
    )
  )

@@

# Create the bar chart
ggplot(total_spent_df, aes(x = contract_size, y = total_amount_billion, fill = contract_size)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(round(total_amount_billion, 2), "B")), 
            vjust = -0.5, 
            size = 5) +  # Adds labels above the bars
  scale_fill_manual(
    values = c("Small Contracts (≤ $100,000)" = "skyblue", "Big Contracts (> $100,000)" = "skyblue")
  ) +
  scale_y_continuous(
    labels = scales::number_format(scale = 1, suffix = "B"),  # Displays y-axis in billions
    expand = expansion(mult = c(0, 0.1))  # Adds space above the bars for labels
  ) +
  labs(
    x = "Contract Size",
    y = "Total Amount Spent (Billions)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

@@

The model diagnostics highlight the limitations of these two linear models, see Appendix TODO for relevant diagnostic plots. The Residuals Vs. Fitted plots indicate large clumps in the first quadrant. This indicates a significant bias in the predictions for both models. The QQ-Normal plots indicate that the majority of the residuals follow a normal distribution. However, there is a significant tail off in the end likely due to a large number of outliers which conclusively violates the normality assumption. The Scale-Location plot for the General Model indicates a funnel shape near the origin, indicates the assumption of homoscedasticity is violated. While the National Defence Model has a clump there is no clear pattern so the assumption of homoscedasticity maybe intact. For all predictors across both models their VIF values are very close to 1, indicating no multicollinearity. Both Cook's distance plots show there is a significant number of outliers in both models.

@@

```{r fig-resid-model}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Residual Vs Fitted for General Model

plot(model, which = 1)

```{r fig-resid-mil}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Residual Vs Fitted for Military Model

plot(mil_model, which = 1)
```

```{r fig-qq-model}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: QQ-Normal General Model

# Q-Q Plot for full model
plot(model, which = 2)

```{r fig-qq-mil}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: QQ-Normal National Defence Model

# Q-Q Plot for National Defence model
plot(mil_model, which = 2)
```

```{r fig-sl-model}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Scale-Location Plot for General model

# Scale-Location Plot for full model
plot(model, which = 3)

```{r fig-sl-mil}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Scale-Location Plot for National Defence model

# Scale-Location Plot for National Defence model
plot(mil_model, which = 3)
```

```{r fig-rl-model}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Residuals vs Leverage for Full model

# Residuals vs Leverage for full model
plot(model, which = 5)
```

```{r fig-rl-mil}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Residuals vs Leverage for National Defence model

# Residuals vs Leverage for National Defence model
plot(mil_model, which = 5)
```

```{r fig-cooks-model}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Cook's Distance for full model

#Identify influential data points using Cook's Distance.
# Cook's Distance for full model
plot(model, which = 4)
```

```{r fig-cooks-mil}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: Cook's Distance for National Defence model

#Identify influential data points using Cook's Distance.
# Cook's Distance for National Defence model
plot(mil_model, which = 4)
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
#| fig-cap: VIF Table

vif_model <- vif(model)
vif_mil_model <- vif(mil_model)

# Create a data frame with VIF values
vif_table <- data.frame(
  Predictor = names(vif_model),
  Model = round(as.numeric(vif_model), 6),
  Mil_Model = round(as.numeric(vif_mil_model), 6)
)

predictor_names <- c(
  "baseline_days" = "Baseline Days",
  "duration_days" = "Duration Days"
)

vif_table$Predictor <- predictor_names[vif_table$Predictor]

# Display the table using knitr's kable for better formatting
kable(vif_table, ,
      col.names = c("Predictor", "General Model", "National Defence Model"),
      align = 'c')
